Subset coercions in \Coq

### 2006-09-14 papers/article/main.tex
Submitted version

\documentclass{llncs}
\usepackage{abbrevs}
\usepackage{array}
\usepackage{alltt}
\usepackage{color}
\def\text{\mbox}
\pagestyle{plain}
%\usepackage{amsmath}
%\usepackage{amsthm}

%\usepackage{qzapfcha}
%\usepackage{cmbright}
%\usepackage{concmath}
\usepackage{ifthen}
\usepackage{url}
\usepackage{varioref}

@@

%\usepackage{framed}
%\usepackage{multicol}

@@

Finally, we can state the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing]
  $`A `G~t, `E T, `G \typea t : T$ and $`A `G~T~U~s, `G \typea T \suba U :
  s$ are decidable problems.
\end{theorem}
\begin{proof}
  Rules are syntax-directed.
\end{proof}

@@

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgement which builds a coercion given two types (we denote it $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
better readability. The judgement $\subimpl{`G}{c}{T}{U}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{T}{`G}$ to an object of type $\ip{U}{`G}$
in context $\ipG{`G}$. We use the algorithmic types to drive the
derivation but the resulting object will be a well-typed \Coq{} term
(lemma \ref{lemma:coercion-correct}). 
% We
% could also directly use the interpreted types but that complicates the
% proof.

@@

As conversion will be preserved by the interpretation $\ip{}{}$,
the rule \irule{SubConv} derives an empty coercion, as the target system
will be able to derive itself that $\ip{T}{`G} `= \ip{U}{`G}$.
Similarly for \irule{SubHnf}, we rely on the target's system conversion
rule so that the coercion $c$ of codomain $\ip{\hnf{U}}{`G}$ can be seen
as an object of type $\ip{U}{`G}$.

@@

It must be stressed that we do not lose any information when
generating obligations as we have the whole term context at hand,
avoiding the complicated proof obligation generation of \PVS.
Here we would have $l : \mysubset{l}{\texttt{list}~A}{l \neq []}, Heql : \eltpit~l =
[]$. We have tactics that clean the goal and context (deconstructing
subsets and simplifying to present $l : \texttt{list}~A, H : l \neq [], Heql : l = []$
to the user) but it does not match the usability of \PVS yet.

@@

\subsection{Related work}
Other efforts to build certified programs using type theory include
Ca\-yenne \cite{Augustsson99} and Epigram
\cite{DBLP:journals/jfp/McBrideM04}. Cayenne offers dependent
types and general recursion at the expense of a non-terminating type
checker; it can be thought of as a testbed for developing languages with
dependent types but is not so much aimed at building certified programs.
Epigram ought to become a complete programming language with dependent
types at its heart. Instead of using a phase distinction to separate
coding and proving, Epigram is based on an interactive, 2-dimensional editing process
where code and proofs are written incrementally to get a complete program.
Using type annotations and with the help of the editor for structuring, 
one is able to write programs with precis e specifications. However it
is not obvious if it could be made usable by programmers because of the
paradigm change and scaling issues. DML \cite{Xi98:PhD} is more akin to
comparison with our solution. It provides a way to use a restricted set
of dependent types in \ML programs, carefully chosen so as to keep
type-checking decidable with the help of an automatic prover. Our method
subsumes this one as it is perfectly possible to integrate automatic
proof tools with \Coq to discharge the generated obligations.

@@

\paragraph{Acknowledgements}
I'd like to thank Christine Paulin-Mohring for directing this work.
\bibliography{my,bib-joehurd,barras,pvs-bib,bcp,Luo,biblio-subset,%
cparent,demons,demons2,demons3}
\bibliographystyle{splncs}
% \renewcommand{\thefootnote}{}
% \footnotetext{This article was typeset in \LaTeX~using the
%   \texttt{Concrete Math} font}

>>>>

\documentclass{llncs}
\usepackage{abbrevs}
\usepackage{array}
\usepackage{alltt}
\usepackage{color}
\def\text{\mbox}
\pagestyle{plain}
\usepackage{ifthen}
\usepackage{url}
\usepackage{varioref}

@@

@@

Finally, we can state the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing]
  $`G \typea t : T$ and $`G \typea T \suba U :
  s$ are decidable problems.
\end{theorem}
\begin{proof}
  Rules are syntax-directed.
\end{proof}

@@

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgement which builds a coercion given two types (denote by $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
better readability. The judgement $\subimpl{`G}{c}{T}{U}{s}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{T}{`G}$ to an object of type $\ip{U}{`G}$
in context $\ipG{`G}$. We use the algorithmic types to drive the
derivation but the resulting object will be a well-typed \Coq{} term
(lemma \ref{lemma:coercion-correct}). 
% We
% could also directly use the interpreted types but that complicates the
% proof.

@@

As conversion will be preserved by the interpretation $\ip{}{}$,
the rule \irule{SubConv} derives an empty coercion, as the target system
will be able to derive itself that $\ip{T}{`G} `= \ip{U}{`G}$.
Similarly for \irule{SubHnf}, we rely on the target system's conversion
rule so that the coercion $c$ of codomain $\ip{\hnf{U}}{`G}$ can be seen
as an object of type $\ip{U}{`G}$.

@@

It must be stressed that we do not lose any information when
generating obligations as we have the whole term context at hand,
avoiding the complicated proof obligation generation of \PVS \cite{PVS-Semantics:TR}.
Here we would have $l : \mysubset{l}{\texttt{list}~A}{l \neq []}, Heql : \eltpit~l =
[]$. We have tactics that clean the goal and context (deconstructing
subsets and simplifying to present $l : \texttt{list}~A, H : l \neq [], Heql : l = []$
to the user) but it does not match the usability of \PVS yet.

@@

\subsection{Related work}
Other efforts to build certified programs using type theory include
\Cayenne \cite{Augustsson99} and \Epigram
\cite{DBLP:journals/jfp/McBrideM04}. \Cayenne offers dependent
types and general recursion at the expense of a non-terminating type
checker; it can be thought of as a testbed for developing languages with
dependent types but is not so much aimed at building certified programs.
\Epigram ought to become a complete programming language with dependent
types at its heart. Instead of using a phase distinction to separate
coding and proving, \Epigram is based on an interactive, 2-dimensional editing process
where code and proofs are written incrementally to get a complete program.
Using type annotations and with the help of the editor for structuring, 
one is able to write programs with precis e specifications. However it
is not obvious if it could be made usable by programmers because of the
paradigm change and scaling issues. The \DML language \cite{Xi98:PhD} is more akin to
comparison with our solution. It provides a way to use a restricted set
of dependent types in \ML programs, carefully chosen so as to keep
type-checking decidable with the help of an automatic prover. Our method
should subsume this one as it is perfectly possible to integrate automatic
proof tools with \Coq to discharge the generated obligations.

@@

\paragraph{Acknowledgements}
I would like to thank Christine Paulin-Mohring for directing this work and
Jean-Christophe Filli\^atre for discussions on previous versions of this paper.
\bibliography{my,bib-joehurd,barras,pvs-bib,bcp,Luo,biblio-subset,%
cparent,demons,demons2,demons3}
\bibliographystyle{splncs}
% \renewcommand{\thefootnote}{}
% \footnotetext{This article was typeset in \LaTeX~using the
%   \texttt{Concrete Math} font}
